Evolution of vertical and oblique transmission under fluctuating selection

### 2017-12-11 ms/PNAS/submission.tex
pnas branch, middle of revision

\begin{abstract}
Conditions for the evolution and maintenance of social learning, in competition with individual learning, under fluctuating selection, is a well-studied problem in the theory of cultural evolution.
Here we study competition between vertical and oblique cultural transmission of a dichotomous phenotype under constant, periodically cycling, and randomly fluctuating selection.
Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime.
Under such a selection regime, the fate of a genetic modifier of the rate of vertical transmission depends on the length of the cycle and the strength of selection.
In general, the evolutionarily stable rate of vertical transmission differs markedly from the rate that maximizes the geometric mean fitness of the population.
The evolution of rules of transmission has dramatically different dynamics from the more frequently studied modifiers of recombination, mutation, or migration.
\end{abstract}

@@

In some situations, oblique transmission of biological material is possible.
In bacteria, phenotypes might be determined by heritable mobile genetic elements such as phages \cite{zinder1952genetic}, plasmids \cite{lederberg1946gene}, integrons \cite{mazel2006integrons}, and transposons \cite{salyers2004conjugative}.
Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation \cite{milkman1990molecular}.
In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the mechanism for oblique transmission evolved to facilitate some other function \cite{redfield1993evolution}.

@@

The linear approximation of $F(x)$ near $x^*=0$ (\eqref{11}) does not depend on the order in which phenotypes $A$ and $B$ are favored within a cycle of $k+l$ generations.
Therefore, the local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations, $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle. 
When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria. Supplementary Figure S1 illustrates the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.

@@

% Figure 1
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/A1B2_polymorphism_vert.pdf}
\caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
\textbf{(B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stochastic_env_x_t.pdf}
\caption{Stochastic local stability. 
The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 28).
The fitnesses of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with probability \(1-p\).
The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
According to Result 6, between these lines fixation of either phenotype is not stochastically locally stable, and we expect a stationary polymorphism between the lines.
Here, initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
\end{figure}

%Figure 3
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stoch_two_iid_rvs.pdf}
\caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment. Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitnesses of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5, the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
but the variance also increases.
The fitnesses of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
\end{figure}

@@

Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} \cite{feldman1986evolutionary,altenberg2017unified} to invasion by allele $M$.
A constant environment always leads to fixation of the favored type, independent of $\rho$. We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case). Specifically, from \eqref{32} with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
\begin{equation}
\underline x''=T_2(T_1\underline x),
\label{34}\end{equation}
where the non-linear transformation $\underline x'=T_1\underline x$ is given by \eqref{32} with $w_A=W$, $w_B=w$, and the non-linear transformation $\underline x''=T_2\underline x'$ is given by \eqref{32} with $w_A=w$, $w_B=W$. Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.

@@

 %Figure 4
\begin{figure*}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/{A1B1_modifier_invasions_w0.5}.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate in selection regime \emph{A1B1}.
The figure shows results of numerical simulations of evolution with two modifier alleles (\eqref{32}).
When a modifier allele fixes (frequency\textgreater{}99.9\%), a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines). \textbf{(A)} The frequency of phenotype \emph{A} in the population over time. \textbf{(B)} The frequency of the invading modifier allele over time. \textbf{(C)} The population geometric mean fitness over time; insets zoom in to show that the mean fitness increases slightly with each invasion.
Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
Parameters: vertical transmission rate of the initial resident modifier allele, \(\rho_0 =0.1\); fitness values: $W=1$; $w=0.5$. The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.
See Figure S12 for $w=0.1$ and $0.9$.}
\label{fig:A1B1_modifier_invasions}
\end{figure*}

%Figure 5
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/stoch_modifier_invasions.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
The figure shows results of numerical simulations of evolution with two modifier alleles
(\eqref{32}).
When a modifier allele fixes (frequency\textgreater{}99.9\%),
a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines).
\textbf{(A)} The frequency of phenotype \emph{A} in the population over time.
\textbf{(B)} The frequency of the invading modifier allele over time.
Invading alleles are introduced at frequency 0.01\%; whenever their frequency drops below 0.01\% they are re-introduced.
Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the ratio of fitness values  is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.}\label{fig:stoch_modifier_invasions}
\end{figure}

@@

Values of the evolutionarily stable vertical transmission rate, $\rho^*$, for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$.
Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
$AkBk$ results are plotted in Figure 6B.
In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested.
However, for $AkBk$ regimes with $k>2$, we find $\rho^*\ne 0$, and depending on $w$, $\rho^*$ can be as high as 0.95.
In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained.
This is why, in Figure 6B, no $\rho^*$ points are shown for $AkBk$ with $k>19$.
In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.

@@

Under fluctuating selection, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics \cite{haldane1963polymorphism,karlin1974random,wakano2004evolution}. For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough \cite{liberman2011evolution}. We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection?
First consider the $A1B1$ selection regime. We have the following result.
\medskip

@@

Figure 4 illustrates the increase over time of the geometric mean fitness with decreasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime.
The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.
Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness increases as $\rho$ decreases.
At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.
In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B.
Also in Figure 6A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.
More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.

\begin{table}[h]
\centering
\caption{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
\begin{tabular}{rrcll}
$k$ & $l$ & $w^{\dagger}$ & $\rho^*$ & $\hat\rho$ \\
\midrule
 1	&	1	&	0.1		&	0.000000	&	0.000000	\cr
 1	&	1	&	0.5		&	0.000000	&	0.000000	\cr
 1	&	1	&	0.9		&	0.000000	&	0.000000	\cr
 1	&	2	&	0.1		&	0.000000	&	0.00065		\cr
 1	&	2	&	0.5		&	fixation	&	fixation	\cr
 1	&	2	&	0.9		&	fixation	&	fixation	\cr
 2	&	2	&	0.1		&	0.000000	&	0.000000	\cr
 2	&	2	&	0.5		&	0.000000	&	0.000000	\cr
 2	&	2	&	0.9		&	0.000000	&	0.000000	\cr
 3	&	10	&	0.1		&	>0.4265		&	0.00031		\cr
 3	&	10	&	0.5		&	fixation	&	fixation	\cr
 3	&	10	&	0.9		&	fixation	&	fixation	\cr
 5	&	30	&	0.1		&	>0.1489		&	0.00027     \cr
 5	&	30	&	0.5		&	fixation	&	fixation	\cr
 5	&	30	&	0.9		&	fixation	&	fixation	\cr
 12	&	12	&	0.1	    &	0.84924		&	0.24347		\cr
 12	&	12	&	0.5		&	0.91209		&	0.000000	\cr
 12	&	12	&	0.9		&	0.95686		&	0.000000	\cr
 20	&	20	&	0.1	    &				&	0.223925	\cr
 20	&	20	&	0.5		&	0.94643		&	0.000000	\cr
 20	&	20	&	0.9		&	0.98304		&	0.000000	\cr
 30	&	30	&	0.1	    &				&	0.193280	\cr
 30	&	30	&	0.5		&	0.96331		&	0.000000	\cr
 30	&	30	&	0.9		&	0.99136		&	0.000000	\cr
 50	&	50	&	0.1	    &				&	0.15419		\cr
 50	&	50	&	0.5		&	0.9768		&	0.22107		\cr
 50	&	50	&	0.9		&	0.99581		&	0.000000	\cr
\bottomrule
\end{tabular}

@@

%Figure 6
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../../figures/AkBk_stable_optimal_rate.pdf}
\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig. S5 and S6 for details on the abrupt transition).
\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al. 2011). The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
In all cases, $W=1$. See \emph{Supplemental Material SP6} for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
\end{figure}

@@

 In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model. Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$. Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see \eqref{2}),
 \begin{equation}
 x' = \rho {w_A \over \overline w} x +(1-\rho)x.
 \label{47}\end{equation}
 Then, according to the Wright-Fisher model \cite{ewens2004mathematical}, $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
 \begin{equation}
 P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},
 \label{48}\end{equation}
 for $j=0,1,2,\dots,N$.
 Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain  where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.

>>>>

\begin{abstract}
Vertical and oblique cultural transmission of a dichotomous phenotype is studied under constant, periodic cycling, and randomly fluctuating selection. Conditions are derived for the existence of a stable polymorphism in a periodically cycling selection regime. Under such a selection regime, the fate of a genetic modifier of the rate of vertical transmission depends on the length of the cycle and the strength of selection. In general, the evolutionarily stable rate of vertical transmission differs markedly from the rate that maximizes the geometric mean fitness of the population. The evolution of rules of transmission has dramatically different dynamics from the more frequently studied modifiers of recombination, mutation, or migration.
\end{abstract}

@@

In some situations, oblique transmission of biological material is possible. In bacteria, phenotypes might be determined by heritable mobile genetic elements such as phages \cite{zinder1952genetic}, plasmids \cite{lederberg1946gene}, integrons \cite{mazel2006integrons}, and transposons \cite{salyers2004conjugative}. Similarly, some phenotypes are determined by genes that are commonly converted by uptake of foreign DNA, i.e., transformation \cite{milkman1990molecular}. In these cases, inheritance of a phenotype may combine vertical transmission from the parent cell, and oblique transmission from other cells, even if the latter did not originally evolve for that purpose \cite{redfield1993evolution}.

@@

The local stability properties of the two fixations depend only on the fact that in a cycle of $(k+l)$ generations $A$ is favored $k$ times and $B$ is favored $l$ times, and not their order in the cycle. 
When neither fixation in $A$ or $B$ is stable, there is a protected polymorphism, and we expect to have one or more polymorphic equilibria. Supplementary Figure S1 illustrates the relationship between $k,l$ and $\rho$ that gives polymorphism of $A$ and $B$, or fixation, for different values of $s=W-w$.

@@

\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/A1B2_polymorphism.pdf}
\caption{Stable frequency of phenotype \emph{A} and geometric mean fitness in selection regime
$A1B2$ as a function of the vertical transmission rate \(\rho\) and the fitness of the disfavored phenotype \(w\).
\textbf{(A)} Stable frequency of phenotype $A$ at the end of each three generations cycle.
\textbf{(B)} Geometric average of the stable population mean fitness over the three generations cycle: $(\bar{w}^* \cdot \bar{w}^{**} \cdot \bar{w}^{***})^{1/3}$.
Gray contour lines join \(\rho\) and \(w\) combinations that result in the same stable value.
In all cases, \emph{W=1}.}\label{fig:A1B2_polymorphism}
%\label{fig:frog}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/stochastic_env_x_t.pdf}
\caption{Stochastic local stability. 
The figure shows the frequency of phenotype \emph{A} after \(10^6\) generations in a very large population evolving in a stochastic environment (eq.\ 28).
The fitnesses of phenotypes \emph{A} and \emph{B} are \(w_A=1+s_t\) and \(w_B=1\),
where \(s_t\) is \emph{s} with probability \emph{p} and \emph{-s} with probability \(1-p\).
The gray lines mark combinations of \emph{p} and \emph{s} for which \(E[\log{(1-\rho+\rho\frac{w_A}{w_B})}]=0\) and \(E[\log{(1-\rho+\rho\frac{w_B}{w_A})}]=0\).
According to Result 6, between these lines fixation of either phenotype is not stochastically locally stable, and we expect a stationary polymorphism between the lines.
Here, initial frequency of \(A\) is \(x_0=1/10,000\) and the vertical transmission rate is \(\rho=0.1\).}\label{fig:stochastic_env_x_t}
\end{figure}

%Figure 3
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/stoch_two_iid_rvs.pdf}
\caption{Effect of vertical transmission rate $\rho$ on phenotype polymorphism in a randomly changing environment. Dynamics of the frequency of phenotype \(A\) over time starting at \(x_0=10^{-5}\) when the fitnesses of phenotypes \(A\) and \(B\) are identically and independently distributed random variables.
As the vertical transmission rate \(\rho\) increases from 0.001 to 0.5, the frequency reaches a polymorphic distribution with \(E(x_t)\to0.5\) faster,
but the variance also increases.
The fitnesses of phenotypes \(A\) and \(B\), \(w_A\) and \(w_B\), are both exponential random variables with expected value 2.}\label{fig:stoch_two_iid_rvs}
\end{figure}

@@

Starting with a stable equilibrium where only the $m$ allele is present, we check its {\sl external stability} \cite{feldman1986evolutionary,altenberg2017unified} to invasion by allele $M$.
Constant environment always leads to fixation of the favored type, independent of $\rho$. We therefore assume changing environments, and, in particular, the simple case of the $A1B1$ cycling environment, where a unique stable polymorphism exists and depends on $\rho$ (but see Supplementary Material SP6 for computational analysis of the general $AkBl$ case). Specifically, from \eqref{32} with $w_A =W$, $w_B =w$ in the first generation and $w_A=w$, $w_B=W$ in the second generation, after two generations we have
\begin{equation}
\underline x''=T_2(T_1\underline x),
\label{34}\end{equation}
where $\underline x'=T_1\underline x$ is given by \eqref{32} with $w_A=W$, $w_B=w$, and $\underline x''=T_2\underline x'$ is given by [32] with $w_A=w$, $w_B=W$. Here $\underline x$, $\underline x'$, $\underline x''$ are the frequency vectors.

@@

 %Figure 4
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/A1B1_modifier_invasions.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical
transmission rate in selection regime \emph{A1B1}. The figure shows
results of numerical simulations of evolution with two modifier alleles
(\eqref{32}). When a modifier allele fixes
(frequency\textgreater{}99.9\%), a new modifier allele is introduced
with a vertical transmission rate one order of magnitude lower (vertical
dashed lines). \textbf{(A,D,G)} The frequency of phenotype \emph{A} in
the population over time. \textbf{(B,E,H)} The frequency of the invading
modifier allele over time. \textbf{(C,F,I)} The population geometric mean
fitness over time; insets zoom in to show that the mean fitness 
decreases slightly with each invasion. Parameters: vertical transmission rate of
the initial resident modifier allele, \(\rho_0 =0.1\); fitness values:
$W=1$; $w=0.1$ (\textbf{A-C}), 0.5 (\textbf{D-F}), and 0.9
(\textbf{G-I}). The x-axis is on a log-scale, as each sequential invasion
takes an order of magnitude longer to
complete.}\label{fig:A1B1_modifier_invasions}
\end{figure}

%Figure 5
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/stoch_modifier_invasions.pdf}
\caption{Consecutive fixation of modifiers that reduce the vertical transmission rate $\rho$ under symmetric randomly changing selection.
The figure shows results of numerical simulations of evolution with two modifier alleles
(\eqref{32}).
When a modifier allele fixes (frequency\textgreater{}99.9\%),
a new modifier allele is introduced with a vertical transmission rate one order of magnitude lower (vertical dashed lines).
\textbf{(A)} The frequency of phenotype \emph{A} in the population over time.
\textbf{(B)} The frequency of the invading modifier allele over time.
Parameters: vertical transmission rate of the initial resident modifier allele is \(\rho_0 =0.1\) and the ratio of fitness values  is $w_A/w_B=10$ with probability 0.5 and $w_A/w_B=0.1$ also with probability 0.5.
The x-axis is on a log-scale, as each sequential invasion takes an order of magnitude longer to complete.}\label{fig:stoch_modifier_invasions}
\end{figure}

@@

Values of $\rho^*$ for some $AkBl$ examples (see Supplementary Material SP7 and  Figure S10 for analytical details) are recorded in Table 1 for different values of $w$ relative to $W=1$. Interestingly, for the $A1B2$ regime, with $w=0.1$, the evolutionarily stable value of $\rho$ is substantially greater than zero, as it is in the $A3B10$ and $A5B30$ regimes.
$AkBk$ results are plotted in Figure 6B. In the $A2B2$ regime, $\rho^* =0$ and there is reduction of vertical transmission for all selection values tested. However, for $AkBk$ regimes with $k>2$, we find $\rho^*\ne 0$, and depending on $w$, $\rho^*$ can be as high as 0.95. In Table 1, blank values for $\rho^*$ indicate that our method was numerically unstable  and that a precise value for $\rho^*$ could not be obtained. This is why, in Figure 6B, no $\rho^*$ points are shown for $AkBk$ with $k>19$. In Table 1, the word ``fixation'' indicates that fixation of $B$ occurs, at which point there can be no effect of modification of $\rho$; $\rho^*$ cannot be calculated in such cases.

@@

Under fluctuating selection, the geometric mean fitness of genotypes has been shown to determine their evolutionary dynamics \cite{haldane1963polymorphism,karlin1974random,wakano2004evolution}. For the evolution of mutation rates that are controlled by genetic modifiers, the stable mutation rate and the mutation rate that maximizes the geometric mean fitness of the population appear to be the same when the period of environmental fluctuation is low enough \cite{liberman2011evolution}. We can ask the same question here: is the stable rate $\rho^*$ the same as the rate $\hat\rho$ that maximizes the equilibrium value of the geometric mean fitness under fluctuating selection. First consider the $A1B1$ selection regime. We have the following result.
\medskip

@@

  Figure 4 illustrates the decrease over time of the geometric mean fitness with increasing $\rho$ at a polymorphic equilibrium in the $A1B1$ regime. The values of $\hat\rho$ and $\rho^*$ are the same in $A1B1$ and $A2B2$ regimes, namely both are zero.  Figure 1B shows the geometric mean fitness in the $A1B2$ regime, and we see that for small values of $w$, this mean fitness decreases as $\rho$ increases. At $w=0.1$, Table 1 shows that $\hat\rho$ is 0.00065, whereas $\rho^*=0.821$.  In all $AkBk$ regimes that we tested with $w=0.9$, the value of $\hat\rho$ was also zero, substantially different from the values of $\rho^*$, as shown in Figures 6A and 6B. Also in Figure 6A, we see that with $w=0.1$, $\hat\rho$ changes from zero to positive in the $AkBk$ regimes with $k \ge 12$, while with $w=0.5$, the change occurs at $k=31$.
 In Figure 6, with $w=0.1$, $\hat\rho$ is between 0.15 and 0.24 for $12\le k\le 50$, while with $w=0.5$, $\hat\rho$ exceeds 0.2 for $31\le k\le 50$.  More details on the mismatch between $\rho^*$, which cannot be invaded, and $\hat\rho$, which maximizes geometric mean fitness, are given in Table 1, Supplementary Figures S7 and S8, and Supplemental Material SP7.

\begin{table}%[h]
\centering
\caption{Values of $\rho^*$ (stable $\rho$)$^{\hbox{\sevenrm (a)}}$ and $\hat\rho$ (``optimal'' $\rho$)$^{\hbox{\sevenrm (b)}}$}
\begin{tabular}{rrcll}
$k$&$l$&$w^{\dagger}$&$\hat\rho$&$\rho^*$ \\
\midrule
1&1&0.1&0.0000000&0.0000000\cr
 1&1&0.5&0.0000000&0.0000000\cr
 1&1&0.9&0.0000000&0.0000000\cr
 1&2&0.1&0.00065&0.821\cr
 1&2&0.5&fixation&fixation\cr
 1&2&0.9&fixation&fixation\cr
 2&2&0.1&0.0000000&0.0000000\cr
 2&2&0.5&0.0000000&0.0000000\cr
 2&2&0.9&0.0000000&0.0000000\cr
 3&10&0.1&0.00031&0.619\cr
 3&10&0.5&fixation&fixation\cr
 3&10&0.9&fixation&fixation\cr
 5&30&0.1&0.00027&0.6503\cr
 5&30&0.5&fixation&fixation\cr
 5&30&0.9&fixation&fixation\cr
 12&12&0.1&0.24347&0.84924\cr
 12&12&0.5&0.0000000&0.91209\cr
 12&12&0.9&0.0000000&0.95686\cr
 20&20&0.1&0.223925\cr
 20&20&0.5&0.0000000&0.94643\cr
 20&20&0.9&0.0000000&0.98304\cr
 30&30&0.1&0.1932801\cr
 30&30&0.5&0.0000000&0.96331\cr
 30&30&0.9&0.0000000&0.0000000\cr
 50&50&0.1&0.15419\cr
 50&50&0.5&0.22107&0.9768\cr
 50&50&0.9&0.0000000&0.99581\cr
\bottomrule
\end{tabular}

@@

%Figure 6
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{../pnas_submission/AkBk_stable_optimal_rate.pdf}
\caption{Fitness ``optimal'' and evolutionary stable vertical transmission rate in $AkBk$ selection regime. 
\textbf{(A)} The vertical transmission rate $\hat{\rho}$ that maximized the geometric average of the population mean fitness is zero (complete oblique transmission) when selection cycles quickly between favoring phenotype $A$ and $B$, and then abruptly transitions to $\approx 0.2$, followed by a slow decrease (see Fig. S5 and S6 for details on the abrupt transition).
\textbf{(B)} The evolutionary stable rate $\rho^*$, which cannot be invaded by modifiers with either higher or lower vertical transmission rate $P$, rapidly increases from zero when selection cycles are short ($k=1$ or $2$) to roughly 1 when selection cycles are longer.
The dashed line shows $1-\frac{1}{k-1}$, which fit the values for $w=0.5$ (Carja et al. 2011). The values for $w=0.1$ (blue) could not be calculated for $k > 19$ due to numerical instability when selection is strong and the duration between selection fluctuations is long.
In all cases, $W=1$. See \emph{Supplemental Material SP6} for details on how we calculated the stable rate.}\label{fig:AkBk_stable_optimal_rate}
\end{figure}

@@

 In order to include the effect of random drift due to finite population in the above deterministic model, we use the Wright-Fisher model. Let $X_t$ denote the number of individuals with phenotype $A$ in a population of fixed size $N$ at the $t$-th generation, and suppose $X_t=Nx$. Also, let $x'$ represent the frequency of the phenotype $A$ in the infinite population model in the next generation, namely (see \eqref{2}),
 \begin{equation}
 x'={w_Ax\over\overline w}\rho +(1-\rho)x.
 \label{47}\end{equation}
 Then, according to the Wright-Fisher model \cite{ewens2004mathematical}, $X_{t+1}$, the number of individuals of phenotype $A$ at generation $(t+1)$, is determined by the probability
 \begin{equation}
 P\bigl(X_{t+1} =j\mid X_t =Nx\bigr) =\left(\matrix{N\cr j\cr}\right)\left(x'\right)^j\left(1-x'\right)^{N-j},
 \label{48}\end{equation}
 for $j=0,1,2,\dots,N$.
 Thus the fluctuations in the numbers of phenotypes $A$ and $B$ in the  population of size $N$ are generated by the Wright-Fisher Markov chain  where, given that $X_t=Nx$, $X_{t+1}$ has a binomial distribution with parameters $(N,x')$.

 This Markov chain  has two absorbing states, $X_t =N$ and $X_t=0$, corresponding to the two fixations in $A$ and $B$, respectively, and we are interested in the fixation probabilities and the time to fixation of these two absorbing states as functions of the initial frequency $x$ and also of $\rho$, $w_A$, and $w_B$.
 To these ends we use a diffusion approximation  to the process $\{X_t\}$. Specifically, we compute the mean $\mu(x)$ and the variance $\sigma^2(x)$ of the change in one generation  in the frequency of phenotype $A$ given that at the beginning of the generation $X_t=Nx$.

 To compute $\mu(x)$, observe that by \eqref{47}   
 \begin{equation}
 \an{
 x' -x &={w_Ax\over\overline w}\rho +(1-\rho)x -x =\rho x\left[{w_A\over \overline w}-1\right]\cr
 &= \rho x(1-x){w_A -w_B\over w_Ax +w_B(1-x)},}
 \label{49}\end{equation}
 since $\overline w =w_A x +w_B(1-x)$.
 For the diffusion approximation, it is essential that the differential selection does not have a large effect per individual in each time period  $\Delta t$ $\bigl(\Delta t\simeq{1\over N}\bigr)$. That is, we assume
 \begin{equation}
 w_A -w_B ={s\over N}.
 \label{50}\end{equation}
 Then
 \begin{equation}
 x'-x\simeq {1\over N}\rho sx(1-x)
 \label{51}\end{equation}
 up to terms of order small than ${1\over N}$. Since one generation corresponds to $\Delta t\simeq {1\over N}$, we  conclude that
 \begin{equation}
 \mu(x) =\rho sx(1-x);\quad 0\le x\le 1.
 \label{52}\end{equation}
 In the same way, we can compute $\sigma^2(x)$, which is $x(1-x)$. These values of $\mu(x)$ and $\sigma^2(x)$ allow us to compute $u(x)$, namely
  \begin{equation}
  u(x) ={1-e^{-2\rho sx}\over 1-e^{-2\rho s}}.
 \label{53} \end{equation}
 The expected time to fixation in $A$ starting from a initial frequency of $x$ is is given by
 \begin{equation}
 T(x)={1-u(x)\over\rho s}\int_0^x{e^{2\rho s\xi}-1\over \xi(1-\xi)}d\xi +{u(x)\over\rho s}\int_x^1{1-e^{-2\rho s(1-\xi)}\over \xi(1-\xi)}d\xi,
 \label{54}\end{equation}
 where $u(x)$ is given in \eqref{53}, and in generations, $T(x)$ is multiplied by $N$.
  Unfortunately the integrals in \eqref{54} cannot be done in closed form unless $\rho s=0$, in which case $u(x)=x$ and $T(x)=-2x\ln x -2(1-x)\ln(1-x)$ as Ewens points out on p.\ 160 \cite{ewens2004mathematical}, and only numerical computation of $T(x)$ is possible for specified values of $x$, $\rho$, and $s$.
